---
title: "WHO xMart and Global Health Estimates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{WHO xMart and Global Health Estimates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = '#>')
library(DSIR)
```

DSIR 0.11.0 keeps the legacy GHO OData service as its compatibility default
and supports WHO's public production xMart service through `backend = 'xmart'`.
GHE always uses xMart. Downloads never switch providers after a failure.
Network examples below are shown without running during vignette builds.

## Discover GHO codes and preserve named dimensions

```{r, eval=FALSE}
gho_indicators('mortality', backend = 'xmart')
gho_dimensions('NCDMORT3070', 'DIM_SEX', backend = 'xmart')
ncd <- gho_data('NCDMORT3070', area = 'PHL', year_from = 2020,
                year_to = 2021, dimensions = list(DIM_SEX = 'TOTAL'),
                backend = 'xmart')
gho_clean(ncd, keep_dimensions = TRUE, keep_metadata = TRUE)
```

The public directory identifies a published route and indicator identifier.
Wide tables store named dimensions and numeric measure families; long tables
store explicit dimension type/member pairs. DSIR maps both into its familiar
raw columns. `gho_clean()` keeps 15 columns by default. Optional named fields
preserve category meaning; optional metadata preserves units and a row-level
`provider` column. Raw GHO observations record `Provider` and a
`who_provenance` attribute. Missing labels are resolved with the recorded backend,
even if the session option has changed. Unknown imported origins remain `NA`.
Positions can differ from legacy `Dim1`-`Dim3`. Prefer named filters.

The xMart directory and old catalog cover different codes and publication
vintages. An unknown code warns; it is never replaced by a similar code.
All six GHO query functions accept `backend`. An explicit argument overrides
`DSIR.who_backend` without changing that option; an unset option uses legacy.
Use the same backend for code discovery, dimensions, counts, coverage and data.

```{r, eval=FALSE}
financial <- gho_data('FINANCIALHARDSHIP_PROPORTIONOFPOP', area = 'PHL',
                       backend = 'legacy')
gho_clean(financial, keep_dimensions = TRUE, keep_metadata = TRUE)

# Optional session default; explicit arguments still take precedence.
options(DSIR.who_backend = 'xmart')
gho_indicators('mortality', backend = 'legacy')
```

Successful empty observation selections give an informational message.
`gho_has_data()` returns `FALSE` and `gho_count()` returns `0L` in that case.
Requests or parsing failures warn and return empty tables or `NA`.
An unknown xMart directory code warns. Legacy HTTP 404 responses trigger a
small directory check before an indicator is reported absent. A failed
directory check cannot establish absence. There is no automatic fallback.

## Discover GHE selections

```{r, eval=FALSE}
ghe_dimensions('year')
ghe_dimensions('area')
ghe_dimensions('sex')
ghe_dimensions('age')
ghe_dimensions('measure')
ghe_causes('diabetes')
ghe_causes('stroke')
```

Official year and sex references are small. Geography uses all-cause,
all-age, both-sex observations in the latest year. Age and cause lists use
a small slice for the first available WHO Member State in ISO3 order.
This avoids expensive whole-database grouping. These are discovery lists;
`ghe_coverage()` verifies the actual selection. Some observed age codes have
no label in the published age reference; their labels remain missing.

Cause `level`, `cause_group`, `rankable` and `single_cause` come from the
provider. It publishes no parent-code field; group codes are not parents.
Do not sum a hierarchy total together with its descendants.

## Download selected measures

```{r, eval=FALSE}
phl_deaths <- ghe_data('PHL', year = 2023, age = 'TOTAL', sex = 'TOTAL',
                       cause = 0, measure = 'deaths')
phl_rates <- ghe_data('PHL', year = 2023, age = 'Y40T44',
                      sex = c('FEMALE', 'MALE'), cause = 0,
                      measure = 'death_rate')
ghe_coverage('PHL', age = 'TOTAL', sex = 'TOTAL', cause = 0)
ghe_clean(phl_rates, keep_dimensions = TRUE)
```

Counts are deaths or years, rates are per 100,000, and shares are percentages.
Counts retain published uncertainty bounds. The API supplies no rate/share
bounds: missing bounds are not filled by calculation. `TOTAL` age rates are
crude; a selected five-year group gives an age-specific rate. Age aggregates
can overlap: do not add every returned age group to obtain a total.

`measure = NULL` returns all supported measures as separate rows. The
source-row count therefore differs from the length of the long result.
Coverage counts source observations. All-null filters are rejected; more
than one million source rows require narrower filters or an explicit
`DSIR.who_max_rows` setting. Large filters use a text-body POST internally.

## Reproducible pulls and revised estimates

```{r, eval=FALSE}
phl <- snapshot(
  ghe_data('PHL', year = 2023, age = 'TOTAL', sex = 'TOTAL', cause = 0,
            measure = 'deaths'),
  'data/phl_ghe_2023.rds'
)
attr(phl, 'who_provenance')
```

Reference caches expire after ten minutes and remain in memory. `snapshot()`
writes only to the path you choose. GHE releases can revise earlier years;
the 2023 estimates published by WHO in 2026 are not automatically comparable
with the older 2021 release. DSIR uses `DEX_CMS/GHE_FULL`, not the older
`GHE_FULL_FOCUS` or `GHE_FULL_DD` views.

## WHO 2026 standard population

Beginning with DSIR 0.11.0, `who_std_pop` uses the WHO 2026 Standard Population.
The official table has 18 groups through `85+`. Its rounded percentages sum
to 100.02, so DSIR normalizes them to 100. `std_million` is a proportional
numeric count scaled to one million, derived without integer rounding.

```{r}
who_std_pop
count <- seq_len(nrow(who_std_pop))
population <- rep(10000, nrow(who_std_pop))
age_standardize(count, population, who_std_pop$weight)
```

The former object is replaced. Aggregate observed ages to the published
groups before using its weights; no finer split of `0-4` or `85+` is supplied.
Source: [WHO/HSA/DDA/GHE/2026.4, Table 1, page 6](https://cdn.who.int/media/docs/default-source/gho-documents/global-health-estimates/ghe2023_who_standard_population.pdf).
